Search for Spatially Extended \fermi-LAT Sources Using Two Years of Flight Data

### 2011-08-22 extended_catalog/paper/extended_catalog.tex
better plots

\keywords{
Catalogs;
Fermi Gamma-ray Space Telescope; 
Gamma rays: observations; 
ISM: supernova remnants;
Methods: statistical
pulsar wind nebula
}

\author{
J.~Lande\altaffilmark{3}, 
\altaffiltext{3}{W. W. Hansen Experimental Physics Laboratory, Kavli Institute for Particle Astrophysics and Cosmology, Department of Physics and SLAC National Accelerator Laboratory, Stanford University, Stanford, CA 94305, USA}
}

@@

The Large Area Telescope (LAT) is a pair conversion telescope on the {\em \fermi
Gamma-ray Space Telescope} (\fermi) that been surveying the $\gamma$-ray
sky since 2008 June.  The LAT has broad energy coverage (20 \mev to $>300$
\gev), wide field of view ($\sim 2.4 \sr$), and large effective area
($\sim 8000 \cm^2$ at $>1 \gev$).

Using two years of all-sky surveying data, the LAT Collaboration published
a catalog of 1873 \gev sources \citep{second_cat}. Many of these have been
positionally associated with a variety of source classes including Active
Galactic Nuclei (AGN), pulsars, galaxies, and supernova remnants (SNRs).
While the counterparts of many LAT sources
can be spatially resolved when observed
at other frequencies, detection of an extension at \gev energies by
the LAT is complicated by the large point-spread function (PSF)
of the instrument. The angular resolution for events converting in the
thin conversion foils of the tracker system, defined as the 68\% containment radius for
a point source, is approximately $3.5^{\circ}$ at 100 \mev, improving
to about 0.1$^{\circ}$ at 10 \gev.

However, a number of sources positionally coincident with LAT sources
exhibit extensions at other wavelengths which are comparable to or larger than 
the LAT PSF.  In particular, certain SNRs, pulsar wind nebulae (PWNe),
molecular clouds, normal galaxies, galaxy clusters or dark matter
satellites can therefore be expected to be detected as spatially extended
sources by the LAT.  
The LAT collaboration has previously reported five SNRs which are
spatially extended beyond the LAT PSF at \gev energies 
(IC443, W28, W44,
W51C, and RX J1713.7-3946 \citep{ic443,w28,w44,w51c,rx_j1713_lat}). In
addition, three extended PWNe were detected: MSH 15-52, Vela~X,
and HESS\,J1825-137 \citep{msh1552,velax,fermi_hess_j1825}, as well
as two close-by galaxies, the Large Magellanic and Small Magalenic
Cloud \citep{lmc,smc}, and finally the radio galaxy Centarus A
\citep{cen_a_lat}.

For each of these sources, the extension was determined based on
dedicated analysis of the particular source 
instead of being the outcome of a systematic search of the LAT catalog.
For the systematic scan of all LAT catalog sources presented in this work, we
therefore expect to find additional spatially extended LAT sources, in
particular since many extended sources in the Galactic plane have been
detected at \tev energies using air Cherenkov detectors and \tev and
\gev emission often originates from the same population of high-energy
particles.  A variety of spatially extended sources have been detected
in the Galactic plane at \tev energies. Most prominent was a survey of
the Galactic plane using the High Energy Stereoscopic System (H.E.S.S)
which reported 14 spatially extended sources
 with extensions varying from
$\sim0\fdg1$ to $\sim0\fdg25$ \citep{hess_plane_survey}.
In fact, within
our Galaxy only very few sources (most notably the $\gamma$-ray binaries
LS\,5039 \citep{HESSLS5039}, LS I+61-303 \citep{MAGICLSI, VERITASLSI}
HESS\,J0632+057 \citep{HESS0632}, and the Crab nebula \citep{crab_weekes}
have 
no significant extension detected at \tev energies. Given
these findings, it can be expected that if the source population at \gev
energies has similar properties, that at least some of the \fermi-LAT
sources in our Galaxy should be spatially resolvable at \gev energies.

\hl{
High energy $\gamma$-rays are produced by the decay of $\pi^0$s produced
by hadronic interactions with interstellar matter and by relativistic
electrons due to Inverse Compton (IC) scattering and Bremsstrahlung
radiation.  Many of these spatially extended \tev sources in the
Galactic plane are unidentified. Some of these sources could produce
hadronic cosmic rays up to $\sim 10^{15}$ eV and subsequently radiate
$\gamma$-rays through $\pi^0$ decay \citep{blandford_and_eichler_1987}.
Studying the spectra of these \tev sources at \gev energies may help to
determine the emission mechanisms of these high energy sources.
}

The capability to
spatially resolve GeV $\gamma$-ray sources is
important for several reasons. 
First, multiple potential counterparts detected at other frequencies
exist within the localization confidence region of a typical LAT source
along the Galactic plane.
Finding a coherent
source extension across different energy bands allows a firm association
of a LAT source to an otherwise confused counterpart.  Also, due
to the strong energy dependence of the LAT PSF,
the spatial and spectral information
about a source do not simply decouple. An inaccurate spatial model will bias
the spectral model of the source. Specifically, performing a spectral
analysis on a spatially extended source modeled as point-like will
systematically shift the spectral index to be softer.  Furthermore,
correctly modeling a source's extension is important for improving
the model of the sky and removing excess residuals, for example in the
region around surrounding the Large Magellanic Cloud \citep{first_cat}.
Additionally, such excess residuals potentially bias the significance
and measured spectrum of neighboring sources in the densely populated
Galactic plane.

@@

Morphological studies of sources in the \gev energy range using the
LAT are challenging because of the strongly energy-dependent PSF.
Additional complications arise for sources along the Galactic plane due to
systematic uncertainties in the Galactic diffuse emission.  The LAT's
PSF is limited at its lowest detectable energies by multiple scattering in the tracker
and the 68\% containment radius of the PSF is approximately $5\fdg1$
at 100 \mev  when averaged over instrument acceptance and including
photons which convert in either the thick or thin layers of the
tracker. The PSF improves with energy and approaches a limit given
by the granularity of the silicon strips and is $0\fdg14$ at 10 \gev
\citep{on_orbit_calibration}.  However, since most high energy
astrophysical sources have power-law spectra, the better resolution of the higher
energy photons is offset by their low
statistics. Therefore a detailed investigation has been performed in this
work to determine the optimal energy range to measure source extension
(section \ref{extended_source_search_method}).

@@

This analysis is not feasible using the standard LAT likelihood
analysis tool \gtlike\footnote{\gtlike is distributed publicly by the
Fermi Science Support Center \citep{fssc}} because \gtlike can only 
fit the spectral parameters of the model unless a more sophisticated
iterative procedure is used to also test various source morphologies.
We note that \gtlike has been used in the past in several studies of
source extension in the LAT collaboration \citep{lmc,smc,w28,w51c}.
Based upon a set of \gtlike maximum likelihood fits at fixed extensions,
a profile of the likelihood as a function of the extension was calculated.
This approach is not optimal because the position, extension, and
spectrum of the source must be simultaneously fit to find the best fit
parameters and to correctly compute the statistical significance of the
detection.  Furthermore since the \gtlike likelihood profile approach
is computationally intensive, no large-scale Monte Carlo effort has been
run to validate it.

The approach presented here is based on a second maximum likelihood
fitting package developed in the LAT collaboration called \pointlike
\citep{first_cat,matthew_kerr_thesis}.  We extended the code to allow
a simultaneous fit of the source extension together with the position
and the spectral parameters.

@@

In \pointlike, it is assumed that the spatial and spectral
model of an extended source are separable, i.e. that the source
model $M(l,b,E)=S(l,b)\times X(E)$ where $S(l,b)$ is the spatial
distribution and $X(E)$ is the spectral distribution.  To fit an extended source,
\pointlike convolves the extended source shape with the PSF (as a
function of energy) and uses the \minuit library to maximize the
likelihood by simultaneously varying the position and extension of
the source \citep{minuit_documentation}.  As will be described in
section~\ref{monte_carlo_validation}, simultaneously fitting the position
and extension is important to correctly calculate the statistical
significance of the detection of extension.  For each position and
extension, the spectral parameters of the sky model are refit.  To avoid
projection effects, the fitted source position parameters are not the
source's longitude and latitude but instead the source's displacement
in a rotated reference frame.

The significance of the extension of a source can be calculated
from \tsext, which 
is the test statistic of the likelihood ratio test of
a model with a spatially extended source and a model
with a point-like source and is defined as
\begin{equation}
  \tsext=2\log(\likelihood_\text{ext}/\likelihood_\text{ps}) = 2 (\loglikelihood_\text{ext}-\loglikelihood_\text{ps})
\end{equation}
where \loglikelihood is the log of the Poisson likelihood.
\pointlike calculates \tsext by fitting a source first with a spatially
extended model and then as a point source.  The interpretation
of \tsext in terms of a statistical significance is discussed in
section~\ref{monte_carlo_validation}.

@@

We estimate the error on the extension of a source by fixing the position
of the source and varying the extension until the likelihood has fallen
by \onehalf, corresponding to a $1\sigma$ error.  This method is
shown schematically in figure~\ref{four_plots_ic443}(a) which shows the
change in \loglikelihood when varying the extension of the SNR IC443.
The localization error is calculated by fixing the extension and spectrum
of the source and fitting to the likelihood function a Gaussian as a function
of position.

@@

We define \tsinc as twice the increase in \loglikelihood fitting the
region as two point sources compared to fitting the region as one point
source: 
\begin{equation}
  \tsinc=2\times(\loglikelihood_\text{2pts}-\loglikelihood_\text{ps}).
\end{equation} 
\tsinc can not be directly compared to \tsext to see which
model is more significant because the models are not nested
\citep{statistics_with_care}. Even though the comparison of \tsext with
\tsinc is not a calibrated test, we find the cases $\tsinc \ll \tsext$
or $\tsinc\gg\tsext$ suggestive and we only consider a source to be
extended if $\tsext>\tsinc$.
Similar to the case of extended sources described in section~\ref{gtlike_crosscheck},
the spectrum of the two point sources can be refit using \gtlike.
We quote the spectral values obtained from \gtlike using the best fit
positions found using \pointlike.

@@

We tested two different models for the
surface brightness profile, either a two dimensional Gaussian
\begin{equation}\label{gauss_pdf}
  I(x,y)=\tfrac{1}{2\pi\sigma^2}\exp\left(-(x^2+y^2)/2\sigma^2\right)
\end{equation}
or a uniform disk
\begin{equation}\label{disk_pdf}
  I(x,y)=
  \begin{cases}
    \frac{1}{\pi\sigma^2} & x^2+y^2\le\sigma^2 \\
    0                      & x^2+y^2>\sigma^2
  \end{cases}
\end{equation}
Although these shapes are significantly different, their convolution
with the PSF look similar.  Figure~\ref{compare_disk_gauss}
demonstrates this by showing $I_\text{src}$, the PSF, and the PDF for
a radially symmetric uniform brightness and the Gaussian surface brightness of size 0\fdg5.\footnote{To
allow a valid comparison between Gaussian and disk shaped morphologies,
we define the source size for this test as the radius containing
68\% of the intensity ($\rsixeight$).  $\rsixeight_\text{,Gaussian}=1.51\sigma$
and $\rsixeight_\text{,disk}=0.82\sigma$ where $\sigma$
is defined in equation~\ref{gauss_pdf} and
equation~\ref{disk_pdf} respectively.} This plot shows that the LAT is 
not very sensitive to probing the exact structure of an extended source.
In the search for extended sources, we therefore use only a uniform
disk as our spatial extension shape. We then quote the radius of the
disk edge as the size of the source.

@@

We test the performance of \pointlike fitting the extension of
point-like sources.
\cite{mattox_egret} discuss that the test statistic distribution
for a likelihood ratio test on the existence of a source at
a given position is 
\begin{equation}\label{ts_ext_distribution}
  P(\ts)=\onehalf(\chi^2_1(\ts)+\delta(\ts)).
\end{equation}
The particular form of equation \ref{ts_ext_distribution} is
due to the null hypothesis (source flux $\Phi=0$) residing
on the edge of parameter space and the model hypothesis
adding a single degree of freedom. It is plausible to
expect a similar distribution of the test statistic
in the test for source extension as the same conditions
apply (with the source flux $\Phi$ replaced by the source radius $r$ and
$r<0$ being unphysical).
To validate this claim, we performed a Monte Carlo
study to calculate empirical distributions for $\tsext$ and
compared them to equation~\ref{ts_ext_distribution}.

We simulated point sources with various spectral forms using
\gtobssim\footnote{\gtobssim is distributed by the Fermi Science Support
Center \citep{fssc}.} and fit them with \pointlike using both point
and extended source hypotheses.  The point sources were simulated with a power-law
spectral model with integrated fluxes above 100 \mev ranging from $3\times
10^{-9}$ to $10^{-6}$ \phflux in six discrete steps and spectral
indices ranging from 1.5 to 3 in four discrete steps.  These values
were picked to to represent typical source parameter of LAT-detected
sources. The point sources were simulated on top of an isotropic
background with an integrated flux above 100 \mev of $1.5\times10^{-5}$
taken to be the same as the isotropic spectrum measured by EGRET
\citep{sreekumar_isotropic}.  The Monte Carlo simulation was performed
over a one year observation period using a default rocking profile and a
representative livetime fraction of 0.8.  The reconstruction was performed
using 1-100 \gev photons and the Pass 7\_V6 (P7\_V6) Source Instrument
Response Function (IRFs, \cite{lat_on_orbit_psf}).  For each 
significantly detected point source ($\ts>25$), we used \pointlike
to fit it as an extended source and calculate \tsext.

For each set of spectral parameters, $\sim30,000$ statistically independent
simulations were performed. For the dimmer spectral models, many of the
simulations left the source undetected ($\ts<25$)
and were discarded.  Table~\ref{ts_ext_num_sims}
shows the different spectral models used in our study as well as the
number of simulations.  The cumulative density of \tsext is plotted in
figure~\ref{ts_ext_mc}. The $\chi^2_1/2$ distribution of
equation~\ref{ts_ext_distribution} is overlaid for comparison.

Our Monte Carlo study shows broad agreement between simulations and
equation~\ref{ts_ext_distribution}. Nevertheless, the agreement is not
perfect.  It should be noted that the discrepancy seems to be worst for
bright sources, a likely cause being numerical errors in the convolution
becoming more apparent for sufficiently bright sources.  Other possible
reasons for departure from equation~\ref{ts_ext_distribution} could
be that \pointlike ignores energy dispersion which would change the
PSF shape as a function of energy. We emphasize that many of
the empirical distributions lies to the left of the theoretical curve so
using the theoretical distribution would lead to an underestimate of the
statistical significance of a detection. Therefore, we are confident that
$\sqrt{\tsext}$ can be used as a conservative measure of the statistical
significance of a source's extension and use it in the following analysis.

@@

We used the simulation setup described in
section~\ref{monte_carlo_validation}, but instead simulated extended
sources with a radially symmetric uniform surface
brightness and
simulated over the actual two year time interval of LAT observations.  
For each extension and
spectral index, we selected a flux range which bracketed $\tsext=16$
and performed an extension test for $>100$ independent realizations of
ten flux values in this range.
We calculated $\langle\tsext\rangle=16$ by fitting a line to the flux
and $\tsext$ values in this narrow range.

Figure~\ref{index_sensitivity} shows the threshold for sources of four
spectral indices from 1.5 to 3 and extension varying from $\sigma=0\fdg1$
to $2\fdg0$.  The LAT's flux threshold for a significant detection
of source extension drops quickly with
increasing source size and reaches a minimum around 0\fdg5. 
Figure~\ref{index_sensitivity} shows
the threshold using photons with energies between 100 \mev and 100 \gev
and also using only 1-100\gev photons.
Except for very large ($>1\deg$) sources, our detection threshold is
not substantially improved including photons with energies between 100 \mev and
1 \gev.  This is also demonstrated in figure~\ref{four_plots_ic443}(b)
which shows \tsext for the SNR IC443 computed independently in twelve
energy bins between 100 \mev and 100 \gev. For IC443, which has a
spectral index $\sim2.4$, almost the entire increase in \ts comes
from energies above 1 \gev.  On the other hand, other systematic errors
become increasingly important at low energy. For our extension search,
we therefore use only photons with energies above 1 \gev.

\hl{
Figure~\ref{all_sensitivity} shows the flux threshold
as a function of source extension for different background levels (1x,
10x, and 100x the nominal background), different spectral indices, and
two different energy bands (1-100 \gev and 10-100 \gev).  The detection
threshold is higher for sources in regions of higher background.  For a
fixed 1-100 \gev or 10-100 \gev flux, the LAT's detection threshold
is only weakly depends upon the spectral index.  This effect is most
pronounced when using only photons with energies above 10 \gev.  Overlaid
on figure~\ref{all_sensitivity} are the LAT detected extended sources
that will be discussed in sections~\ref{validate_known}
and \ref{new_ext_srcs_section}.  The extension thresholds are tabulated
in table~\ref{all_sensitivity_table}.

Finally, figure~\ref{time_sensitivity} shows the LAT's projected detection
threshold to extension after 10 years against 10 times the isotropic
background. This background is representative of the background near
the Galactic plane.  For small extended sources, our detection threshold
improves by a factor larger than the square root of time because at high energies, where we
are most sensitive to extension, the background levels are in the Poisson
instead of Gaussian regime.  For large extended sources, the relevant
background is larger and the improvement is closer to the expected
factor of the square root of time.
}

@@

As was shown in section~\ref{extension_sensitivity}, we gain little in sensitivity using photons with energies
below 1 \gev. On the other hand, the large PSF at low energy makes us
more susceptible to systematic errors arising from source confusion due
to multiple point sources and susceptible to incorrect modeling of the
Galactic diffuse emission. In addition, the Galactic diffuse emission
is more pronounced at lower energies due to its steep energy spectrum.
For that reason, we performed our search using photons only between 1
\gev and 100 \gev.

@@

Finally, when analyzing the region, we automatically removed other 2FGL
sources which were within 0\fdg5 of the source of interest. This was
done to avoid a concern that extended sources are included in 2FGL as
multiple point sources. These spurious sources could distort the
extension fit.  Instead, when a source is found to be significantly
extended (i.e. $\tsext>16$), we perform the dual localization procedure
(described in section~\ref{dual_localization_method} to compare the
extended source hypothesis to the hypothesis of two independent point
sources. Only sources with $\tsext>\tsinc$ are considered as extended.

>>>>

\keywords{
Catalogs;
Fermi Gamma-ray Space Telescope; 
Gamma rays: observations; 
ISM: supernova remnants;
Methods: statistical;
pulsar wind nebula
}

\author{
J.~Lande\altaffilmark{2},
S.~Funk\altaffilmark{2},
M.~Ackermann (???),
A.~Allafort\altaffilmark{2},
K.~Bechtol\altaffilmark{2},
M.-H.~Grondin\altaffilmark{53},
\ldots
\altaffiltext{2}{W. W. Hansen Experimental Physics Laboratory, Kavli Institute for Particle Astrophysics and Cosmology, Department of Physics and SLAC National Accelerator Laboratory, Stanford University, Stanford, CA 94305, USA}
\altaffiltext{53}{Institut f\"ur Astronomie und Astrophysik, Universit\"at T\"ubingen, D 72076 T\"ubingen, Germany}
}

@@

The Large Area Telescope (LAT) is a pair conversion telescope on the {\em \fermi
Gamma-ray Space Telescope} (\fermi) that been surveying the $\gamma$-ray
sky since 2008 June.  The LAT has broad energy coverage (20 \mev to $>300$
\gev), wide field of view ($\sim 2.4 \sr$), and large effective area
($\sim 8000\ \cm^2$ at $>1 \gev$).

Using two years of all-sky surveying data, the LAT Collaboration published
a catalog of 1873 \gev sources \citep{second_cat}. Many of these have
been positionally associated with a variety of source classes including
Active Galactic Nuclei (AGN), pulsars, normal galaxies, and supernova
remnants (SNRs).  While the counterparts of many LAT sources can be
spatially resolved when observed at other frequencies, detection of
spatial extension of these sources at \gev energies is difficult because
the size of the point-spread function (PSF) of the LAT is comparable
to the typical size of many of these sources.

However, a number of sources positionally coincident with LAT sources
exhibit extensions at other wavelengths which are comparable to or larger than 
the LAT PSF.  In particular, certain SNRs, pulsar wind nebulae (PWNe),
molecular clouds, normal galaxies, galaxy clusters or dark matter
satellites could be expected to be detected as spatially extended
sources by the LAT.  
The LAT Collaboration has previously reported on five SNRs which are
spatially extended beyond the LAT PSF at \gev energies 
(IC443, W28, W44,
W51C, and RX J1713.7-3946 \citep{ic443,w28,w44,w51c,rx_j1713_lat}). In
addition, three extended PWNe were detected: MSH 15-52, Vela~X,
and HESS\,J1825-137 \citep{msh1552,velax,fermi_hess_j1825}, as well
as two close-by galaxies, the Large and Small Magalenic
Cloud \citep{lmc,smc}, and finally the radio galaxy Centarus A
\citep{cen_a_lat}.
For each of these sources, the extension was determined from a dedicated
analysis of the particular source and so a systematic
scan of all LAT detected sources could uncover
additonal spatially extended sources.

\hl{
Many extended sources in the Galactic plane have been detected at \tev
energies using air Cherenkov detectors.  Most prominent was a survey of
the Galactic plane using the High Energy Stereoscopic System (H.E.S.S)
which reported 14 spatially extended sources with extensions varying from
$\sim0\fdg1$ to $\sim0\fdg25$ \citep{hess_plane_survey}.  In fact, within
our Galaxy only very few sources (most notably the $\gamma$-ray binaries
LS\,5039 \citep{HESSLS5039}, LS I+61-303 \citep{MAGICLSI, VERITASLSI}
HESS\,J0632+057 \citep{HESS0632}, and the Crab nebula \citep{crab_weekes})
have no significant extension detected at \tev energies.  High energy
$\gamma$-rays from these sources are produced by the decay of $\pi^0$s
produced by hadronic interactions with interstellar matter and by
relativistic electrons due to Inverse Compton (IC) scattering and
Bremsstrahlung radiation \citep{blandford_and_eichler_1987}.  It is likely
that the \gev and \tev emission from these sources originates from the
same population of high-energy particles and so at least some of these
sources should be detected as extended sources at \gev energies. By
studying the spectra of these \tev sources at \gev energies, we may help
to determine the emission mechanisms of these high energy sources.
}

The capability to spatially resolve \gev $\gamma$-ray sources is important
for several reasons.  First, multiple potential counterparts detected at
other frequencies exist within the localization confidence region of a
typical LAT source along the Galactic plane.  Finding a coherent source
extension across different energy bands can help to associate a
LAT source to an otherwise confused counterpart.  Also, due to the strong
energy dependence of the LAT PSF, the spatial and spectral information
about a source do not simply decouple. An inaccurate spatial model
will bias the spectral model of the source. Specifically, 
modeling a spatially extended source as point-like will systematically
shift a spectral analysis to softer indices. Furthermore,
correctly modeling a source's extension is important for improving
the model of the sky and removing excess residuals, for example in the
region around surrounding the Large Magellanic Clouds \citep{first_cat}.
Such excess residuals potentially bias the significance
and measured spectrum of neighboring sources in the densely populated
Galactic plane.

@@

Morphological studies of sources using the LAT are challenging
because of the strongly energy-dependent PSF that is comparable in
size to the extension of many sources expected to be detected at
\gev energies.  Additional complications arise for sources along
the Galactic plane due to systematic uncertainties in the Galactic
diffuse emission.  The LAT's PSF is limited at its lowest detectable
energies by multiple scattering in the silicon strip tracking section
of the detector and is several degrees at 100 \mev.  The PSF improves
with energy approaching a 68\% containment radius of $\sim0\fdg2$ at
the highest energies (when averaged over the acceptance of the LAT)
and is limited by the granularity of the silicon strips in the tracker
\citep{atwood_LAT_mission,on_orbit_calibration,lat_on_orbit_psf}.\footnote{More
information about the performance of the LAT can be found at the Fermi
Science Support Center (FSSC,\url{http://fermi.gsfc.nasa.gov}).} However,
since most high energy astrophysical sources have spectra that decrease
rapidly with increasing energy, the improved resolution of the higher
energy photons is offset by their low statistics. Therefore sophisticated
analysis techniques are required to maximize our sensitivity
to these extended sources.

@@

This type of analysis is not feasible using the standard LAT likelihood
analysis tool \gtlike\footnote{\gtlike is distributed publicly by the
FSSC.} because \gtlike can only
fit the spectral parameters of the model unless a more sophisticated
iterative procedure is used to also test various source morphologies.
We note that \gtlike has been used in the past in several studies of
source extension in the LAT Collaboration \citep{lmc,smc,w28,w51c}.
Based upon a set of \gtlike maximum likelihood fits at fixed extensions,
a profile of the likelihood as a function of the extension was calculated.
This approach is not optimal because the position, extension, and
spectrum of the source must be simultaneously fit to find the best fit
parameters and to correctly compute the statistical significance of the
detection.  Furthermore since the \gtlike likelihood profile approach
is computationally intensive, no large-scale Monte Carlo effort has been
run to validate it.

The approach presented here is based on a second maximum likelihood
fitting package developed in the LAT Collaboration called \pointlike
\citep{first_cat,matthew_kerr_thesis}.  We extended the code to allow
a simultaneous fit of the source extension together with the position
and the spectral parameters.

@@

In \pointlike, it is assumed that the spatial and spectral
model of an extended source are separable, i.e. that the source
model $M(l,b,E)=S(l,b)\times X(E)$ where $S(l,b)$ is the spatial
distribution and $X(E)$ is the spectral distribution.  To fit an extended source,
\pointlike convolves the extended source shape with the PSF (as a
function of energy) and uses the \minuit library to maximize the
likelihood by simultaneously varying the position and extension of
the source \citep{minuit_documentation}.  As will be described in
Section~\ref{monte_carlo_validation}, simultaneously fitting the position
and extension is important to correctly calculate the statistical
significance of the detection of extension.  For each position and
extension, the spectral parameters of the sky model are refit.  To avoid
projection effects, the fitted source position parameters are not the
source's longitude and latitude but instead the source's displacement
in a rotated reference frame.

The significance of the extension of a source can be calculated
from \tsext, which 
is the test statistic of the likelihood ratio test of
a model with a spatially extended source and a model
with a point-like source and is defined as
\begin{equation}
  \tsext=2\log(\likelihood_\text{ext}/\likelihood_\text{ps}) = 2 (\loglikelihood_\text{ext}-\loglikelihood_\text{ps})
\end{equation}
where \loglikelihood is the log of the Poisson likelihood.
\pointlike calculates \tsext by fitting a source first with a spatially
extended model and then as a point source.  The interpretation
of \tsext in terms of a statistical significance is discussed in
Section~\ref{monte_carlo_validation}.

@@

We estimate the error on the extension of a source by fixing the position
of the source and varying the extension until the likelihood has fallen
by \onehalf, corresponding to a $1\sigma$ error.  This method is
shown schematically in Figure~\ref{four_plots_ic443}(a) which shows the
change in \loglikelihood when varying the extension of the SNR IC443.
The localization error is calculated by fixing the extension and spectrum
of the source and fitting to the likelihood function a Gaussian as a function
of position.

@@

We define \tsinc as twice the increase in \loglikelihood fitting the
region as two point sources compared to fitting the region as one point
source: 
\begin{equation}
  \tsinc=2\times(\loglikelihood_\text{2pts}-\loglikelihood_\text{ps}).
\end{equation} 
\tsinc can not be directly compared to \tsext to see which
model is more significant because the models are not nested
\citep{statistics_with_care}. Even though the comparison of \tsext with
\tsinc is not a calibrated test, we find the cases $\tsinc \ll \tsext$
or $\tsinc\gg\tsext$ suggestive and we only consider a source to be
extended if $\tsext>\tsinc$.
Similar to the case of extended sources described in Section~\ref{gtlike_crosscheck},
the spectrum of the two point sources can be refit using \gtlike.
We quote the spectral values obtained from \gtlike using the best fit
positions found using \pointlike.

@@

We tested two different models for the
surface brightness profile, either a two dimensional Gaussian
\begin{equation}\label{gauss_pdf}
  I(x,y)=\tfrac{1}{2\pi\sigma^2}\exp\left(-(x^2+y^2)/2\sigma^2\right)
\end{equation}
or a uniform disk
\begin{equation}\label{disk_pdf}
  I(x,y)=
  \begin{cases}
    \frac{1}{\pi\sigma^2} & x^2+y^2\le\sigma^2 \\
    0                      & x^2+y^2>\sigma^2
  \end{cases}
\end{equation}
Although these shapes are significantly different, their convolution
with the PSF look similar.  Figure~\ref{compare_disk_gauss}
demonstrates this by showing $I_\text{src}$, the PSF, and the PDF for
a radially symmetric uniform brightness and the Gaussian surface brightness of size 0\fdg5.\footnote{To
allow a valid comparison between Gaussian and disk shaped morphologies,
we define the source size for this test as the radius containing
68\% of the intensity ($\rsixeight$).  $\rsixeight_\text{,Gaussian}=1.51\sigma$
and $\rsixeight_\text{,disk}=0.82\sigma$ where $\sigma$
is defined in Equation~\ref{gauss_pdf} and
Equation~\ref{disk_pdf} respectively.} This plot shows that the LAT is 
not very sensitive to probing the exact structure of an extended source.
In the search for extended sources, we therefore use only a uniform
disk as our spatial extension shape. We then quote the radius of the
disk edge as the size of the source.

@@

We test the performance of \pointlike fitting the extension of
point-like sources.
\cite{mattox_egret} discuss that the test statistic distribution
for a likelihood ratio test on the existence of a source at
a given position is 
\begin{equation}\label{ts_ext_distribution}
  P(\ts)=\onehalf(\chi^2_1(\ts)+\delta(\ts)).
\end{equation}
The particular form of Equation \ref{ts_ext_distribution} is
due to the null hypothesis (source flux $\Phi=0$) residing
on the edge of parameter space and the model hypothesis
adding a single degree of freedom. It is plausible to
expect a similar distribution of the test statistic
in the test for source extension as the same conditions
apply (with the source flux $\Phi$ replaced by the source radius $r$ and
$r<0$ being unphysical).
To validate this claim, we performed a Monte Carlo
study to calculate empirical distributions for $\tsext$ and
compared them to Equation~\ref{ts_ext_distribution}.

We simulated point sources with various spectral forms using
\gtobssim\footnote{\gtobssim is distributed publicly by the FSSC.} and fit them with \pointlike using both point
and extended source hypotheses.  The point sources were simulated with a power-law
spectral model with integrated fluxes above 100 \mev ranging from $3\times
10^{-9}$ to $10^{-6}$ \phflux in six discrete steps and spectral
indices ranging from 1.5 to 3 in four discrete steps.  These values
were picked to to represent typical source parameter of LAT-detected
sources. The point sources were simulated on top of an isotropic
background with an integrated flux above 100 \mev of $1.5\times10^{-5}$
taken to be the same as the isotropic spectrum measured by EGRET
\citep{sreekumar_isotropic}.  The Monte Carlo simulation was performed
over a one year observation period using a default rocking profile and a
representative livetime fraction of 0.8.  The reconstruction was performed
using 1-100 \gev photons and the Pass 7\_V6 (P7\_V6) Source Instrument
Response Function (IRFs, \cite{lat_on_orbit_psf}).  For each 
significantly detected point source ($\ts>25$), we used \pointlike
to fit it as an extended source and calculate \tsext.

For each set of spectral parameters, $\sim30,000$ statistically independent
simulations were performed. For the dimmer spectral models, many of the
simulations left the source undetected ($\ts<25$)
and were discarded.  Table~\ref{ts_ext_num_sims}
shows the different spectral models used in our study as well as the
number of simulations.  The cumulative density of \tsext is plotted in
Figure~\ref{ts_ext_mc}. The $\chi^2_1/2$ distribution of
Equation~\ref{ts_ext_distribution} is overlaid for comparison.

Our Monte Carlo study shows broad agreement between simulations and
Equation~\ref{ts_ext_distribution}. Nevertheless, the agreement is not
perfect.  It should be noted that the discrepancy seems to be worst for
bright sources, a likely cause being numerical errors in the convolution
becoming more apparent for sufficiently bright sources.  Other possible
reasons for departure from Equation~\ref{ts_ext_distribution} could
be that \pointlike ignores energy dispersion which would change the
PSF shape as a function of energy. We emphasize that many of
the empirical distributions lies to the left of the theoretical curve so
using the theoretical distribution would lead to an underestimate of the
statistical significance of a detection. Therefore, we are confident that
$\sqrt{\tsext}$ can be used as a conservative measure of the statistical
significance of a source's extension and use it in the following analysis.

@@

We used the simulation setup described in
Section~\ref{monte_carlo_validation}, but instead simulated extended
sources with a radially symmetric uniform surface
brightness and
simulated over the actual two year time interval of LAT observations.  
For each extension and
spectral index, we selected a flux range which bracketed $\tsext=16$
and performed an extension test for $>100$ independent realizations of
ten flux values in this range.
We calculated $\langle\tsext\rangle=16$ by fitting a line to the flux
and $\tsext$ values in this narrow range.

Figure~\ref{index_sensitivity} shows the threshold for sources of four
spectral indices from 1.5 to 3 and extension varying from $\sigma=0\fdg1$
to $2\fdg0$.  The LAT's flux threshold for a significant detection
of source extension drops quickly with
increasing source size and reaches a minimum around 0\fdg5. 
Figure~\ref{index_sensitivity} shows
the threshold using photons with energies between 100 \mev and 100 \gev
and also using only 1-100\gev photons.
Except for very large ($>1\deg$) sources, our detection threshold is
not substantially improved including photons with energies between 100 \mev and
1 \gev.  This is also demonstrated in Figure~\ref{four_plots_ic443}(b)
which shows \tsext for the SNR IC443 computed independently in twelve
energy bins between 100 \mev and 100 \gev. For IC443, which has a
spectral index $\sim2.4$, almost the entire increase in \ts comes
from energies above 1 \gev.  On the other hand, other systematic errors
become increasingly important at low energy. For our extension search,
we therefore use only photons with energies above 1 \gev.

\hl{
Figure~\ref{all_sensitivity} shows the flux threshold
as a function of source extension for different background levels (1x,
10x, and 100x the nominal background), different spectral indices, and
two different energy bands (1-100 \gev and 10-100 \gev).  The detection
threshold is higher for sources in regions of higher background.  For a
fixed 1-100 \gev or 10-100 \gev flux, the LAT's detection threshold
is only weakly depends upon the spectral index.  This effect is most
pronounced when using only photons with energies above 10 \gev.  Overlaid
on Figure~\ref{all_sensitivity} are the LAT detected extended sources
that will be discussed in Sections~\ref{validate_known}
and \ref{new_ext_srcs_section}.  The extension thresholds are tabulated
in Table~\ref{all_sensitivity_table}.

Finally, Figure~\ref{time_sensitivity} shows the LAT's projected detection
threshold to extension after 10 years against 10 times the isotropic
background. This background is representative of the background near
the Galactic plane.  For small extended sources, our detection threshold
improves by a factor larger than the square root of time because at high energies, where we
are most sensitive to extension, the background levels are in the Poisson
instead of Gaussian regime.  For large extended sources, the relevant
background is larger and the improvement is closer to the expected
factor of the square root of time.
}

@@

As was shown in Section~\ref{extension_sensitivity}, we gain little in sensitivity using photons with energies
below 1 \gev. On the other hand, the large PSF at low energy makes us
more susceptible to systematic errors arising from source confusion due
to multiple point sources and susceptible to incorrect modeling of the
Galactic diffuse emission. In addition, the Galactic diffuse emission
is more pronounced at lower energies due to its steep energy spectrum.
For that reason, we performed our search using photons only between 1
\gev and 100 \gev.

@@

Finally, when analyzing the region, we automatically removed other 2FGL
sources which were within 0\fdg5 of the source of interest. This was
done to avoid a concern that extended sources are included in 2FGL as
multiple point sources. These spurious sources could distort the
extension fit.  Instead, when a source is found to be significantly
extended (i.e. $\tsext>16$), we perform the dual localization procedure
(described in Section~\ref{dual_localization_method} to compare the
extended source hypothesis to the hypothesis of two independent point
sources. Only sources with $\tsext>\tsinc$ are considered as extended.
